Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-14 ind_files/04_field_eqns.tex
Change filenames to be more sensible

If you know and care about enough mathematics for this to really bother you, see \citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions \citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see \cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all the (rather pedantic) questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function \citep{bressloff_stochastic_2010} and collective motion \citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

\myfig{0.5}{Media/3.3_dirac_deltas.png}{\textbf{Schematic description of a function valued birth-death process.} Consider a population of birds in which individuals have varying beak lengths. \textbf{(A)} Each individual in the population can be described as a Dirac delta mass centered at its beak length. This is because each individual has exactly one fixed beak length, and therefore, can be thought of as a distribution centered at that particular beak length and with zero spread. \textbf{(B)} The population as a whole is thus described as a sum of Dirac masses. $N(t)$ here is the size of the population at time $t$. Birth and death of individuals would correspond to the addition and removal of Dirac masses respectively. Note that if we had a large number of individuals, this distribution begins to look like a continuous distribution.}{fig_infD_pop_description}

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
It is known (only for one-dimensional traits) that we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$.\\
We can now use the same trick as earlier and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.\\
\\
\section{The functional system-size expansion}

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_K = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}
\\
\\
We can now conduct a system-size expansion as before by using a functional `Taylor expansion' of the step operators. Recall that the functional version of the Taylor expansion of a functional $F[\rho]$ about a function $\rho_0$ defined on a domain $\Omega \subseteq \mathbb{R}$ is given by:
\begin{equation*}
    F[\rho_0 + \rho] = F[\rho_0] + \int\limits_{\Omega}\rho(x)\frac{\delta F}{\delta \rho_0(x)}dx + \frac{1}{2!}\int\limits_{\Omega}\int\limits_{\Omega}\rho(x)\rho(y)\frac{\delta^2 F}{\delta \rho_0(x)\delta \rho_0(y)}dxdy + \cdots
\end{equation*}
Since $\Delta^{\pm}_{x}[F[\phi]] = F[\phi \pm \delta_x/K]$, we can Taylor expand the RHS to see that our step operators obey
\begin{align}
    \Delta^{\pm}_{x}[F[\phi]] &= F[\phi] \pm \frac{1}{K}\int\limits_{\mathcal{T}}\frac{\delta F}{\delta \phi(y)}\delta_xdy + \frac{1}{2K^2}\int\limits_{\mathcal{T}}\int\limits_{\mathcal{T}}\frac{\delta^2 F}{\delta \phi(y)\delta \phi(z)}\delta_xdy\delta_xdz+\mathcal{O}(K^{-3})\nonumber\\
    &= F[\phi] \pm \frac{1}{K}\frac{\delta F}{\delta \phi(x)} + \frac{1}{2K^2}\frac{\delta^2 F}{\delta \phi(x)^2}+\mathcal{O}(K^{-3})
    \label{KM_ansatz}
\end{align}
Neglecting terms of $\mathcal{O}(K^{-3})$, we can now substitute \eqref{KM_ansatz} into \eqref{M_equation} to obtain:
\begin{equation*}
\begin{split}
\frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[
    \left(-\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi(x)^2}\right)\{b_K(x|\phi)P(\phi,t)\}\right]dx\\
    +K\int\limits_{\mathcal{T}}\left[\left(\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi^2(x)}\right)\{d_K(x|\phi)P(\phi,t)\}\right]dx
\end{split}
\end{equation*}
Rearranging these terms, we obtain a `functional Fokker-Planck equation':
\begin{equation}
\label{functional_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
    \frac{\delta}{\delta\phi(x)}\{\mathcal{A}^{-}(x|\phi)P(\phi,t)\} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\mathcal{A}^{+}(x|\phi)P(\phi,t)\}\right]dx}
\end{equation}
where
\begin{align*}
   \mathcal{A}^{\pm}(x|\phi) &= b_K(x|\phi)\pm d_K(x|\phi) = \frac{1}{K}\left(b(x|\nu)\pm d(x|\nu)\right)
\end{align*}

This constitutes the `mesoscopic' description. For large (but finite) $K$, equation \eqref{functional_FPE} can be analyzed using a weak noise approximation as before.

\section{The infinite population limit}
We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE' models discussed in chapter \ref{chap_AD}. They are also equivalent to the `oligomorphic dynamics' of \citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology \citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of \citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics.\\
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear. Nevertheless, we will pretend all is well and assume that one can do this, bolstered by the fact that we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, as we show below.

\subsection{Some familiar faces: Kimura-Crow and adaptive dynamics}

The `macroscopic' version of our function-valued process corresponds to quantitative genetics and adaptive dynamics (which is a generalization of evolutionary game theory for quantitative traits). We begin with the deterministic process given by \eqref{deterministic_traj}. We assume that the birth and death functions take the form:

@@

As in chapter \ref{chap_BD}, $ b^{\textrm{(mut)}}(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu b^{\textrm{(mut)}}(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b^{\textrm{(mut)}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} b^{\textrm{(mut)}}(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]}

@@

\frac{d \overline{x}}{dt} &= \int\limits_{\mathcal{T}}xw(x|\psi)p(x,t)dx - \overline{w}(t)\int\limits_{\mathcal{T}}xp(x,t)dx+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}x\left[b^{\textrm{(mut)}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b^{\textrm{(mut)}}(y|\psi)dy\right]dx\nonumber\\

>>>>

If you know and care about enough mathematics for this to really bother you, see~\citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The mathematics for these sorts of processes is an active area of research and is comparatively far from well developed. The mathematically rigorous formulation of the kinds of processes we study here falls in the realm of measure-valued branching processes, and is highly technical and rather inaccessible unless one is already comfortable with advanced measure-theoretic notions~\citep{champagnat_unifying_2006,champagnat_individual_2008}. This means that the existing formalism, while admirable in its generality and mathematical rigor, is rather unusable for most biologists, who do not have formal training in analysis (but see~\cite{week_white_2021} for a very friendly introduction to the major ideas through heuristics). One can, however, make progress if they are willing to take some mathematical leaps of faith and sacrifice rigor for the sake of accessibility and heuristic understanding. I adopt this attitude below and hope that all questions of rigor, well-posedness, existence, etc. will be sorted out by some clever mathematicians in the future. Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, the stochastic process we will formulate $\{\phi^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\phi^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function~\citep{bressloff_stochastic_2010} and collective motion~\citep{o_laighleis_minimal_2018}. In the following sections, I will rely heavily on a heuristic object called the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$

@@

\myfig{0.5}{figures/4.1_dirac_deltas.png}{\textbf{Schematic description of a function valued birth-death process.} Consider a population of birds in which individuals have varying beak lengths. \textbf{(A)} Each individual in the population can be described as a Dirac delta mass centered at its beak length. This is because each individual has exactly one fixed beak length, and therefore, can be thought of as a distribution centered at that particular beak length and with zero spread. \textbf{(B)} The population as a whole is thus described as a sum of Dirac masses. $N(t)$ here is the size of the population at time $t$. Birth and death of individuals would correspond to the addition and removal of Dirac masses respectively. Note that if we had a large number of individuals, this distribution begins to look like a continuous distribution.}{fig_infD_pop_description}

@@

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
It is known (only for one-dimensional traits) that we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$.

We can now use the same trick as in chapter \ref{chap_BD} and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}(\mathcal{T})$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.

\section{The functional system-size expansion}

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_{K}(\mathcal{T}) = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}

We can now conduct a system-size expansion as before by using a functional `Taylor expansion' of the step operators. Recall that the functional version of the Taylor expansion of a functional $F[\rho]$ about a function $\rho_0$ defined on a domain $\Omega \subseteq \mathbb{R}$ is given by:
\begin{equation*}
    F[\rho_0 + \rho] = F[\rho_0] + \int\limits_{\Omega}\rho(x)\frac{\delta F}{\delta \rho_0(x)}dx + \frac{1}{2!}\int\limits_{\Omega}\int\limits_{\Omega}\rho(x)\rho(y)\frac{\delta^2 F}{\delta \rho_0(x)\delta \rho_0(y)}dxdy + \cdots
\end{equation*}
Since $\Delta^{\pm}_{x}[F[\phi]] = F[\phi \pm \delta_x/K]$, we can Taylor expand the RHS to see that our step operators obey
\begin{align}
    \Delta^{\pm}_{x}[F[\phi]] &= F[\phi] \pm \frac{1}{K}\int\limits_{\mathcal{T}}\frac{\delta F}{\delta \phi(y)}\delta_xdy + \frac{1}{2K^2}\int\limits_{\mathcal{T}}\int\limits_{\mathcal{T}}\frac{\delta^2 F}{\delta \phi(y)\delta \phi(z)}\delta_xdy\delta_xdz+\mathcal{O}(K^{-3})\nonumber\\
    &= F[\phi] \pm \frac{1}{K}\frac{\delta F}{\delta \phi(x)} + \frac{1}{2K^2}\frac{\delta^2 F}{\delta \phi(x)^2}+\mathcal{O}(K^{-3})
    \label{KM_ansatz}
\end{align}
Neglecting terms of $\mathcal{O}(K^{-3})$, we can now substitute \eqref{KM_ansatz} into \eqref{M_equation} to obtain:
\begin{equation*}
\begin{split}
\frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[
    \left(-\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi(x)^2}\right)\{b_K(x|\phi)P(\phi,t)\}\right]dx\\
    +K\int\limits_{\mathcal{T}}\left[\left(\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi^2(x)}\right)\{d_K(x|\phi)P(\phi,t)\}\right]dx
\end{split}
\end{equation*}
Rearranging these terms, we obtain a `functional Fokker-Planck equation':
\begin{equation}
\label{functional_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
    \frac{\delta}{\delta\phi(x)}\{\mathcal{A}^{-}(x|\phi)P(\phi,t)\} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\mathcal{A}^{+}(x|\phi)P(\phi,t)\}\right]dx}
\end{equation}
where
\begin{align*}
   \mathcal{A}^{\pm}(x|\phi) &= b_K(x|\phi)\pm d_K(x|\phi) = \frac{1}{K}\left(b(x|\nu)\pm d(x|\nu)\right)
\end{align*}

This constitutes the `mesoscopic' description of our process. We can once again appeal to the link between Fokker-Planck equations and Langevins~\citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) for finite-dimensional stochastic processes is part of the standard mathematical canon, the corresponding equivalence is much less well understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear for general measure spaces - We may need $\mathcal{M}_{K}(\mathcal{T})$ to have a considerable amount of structure (Ex: Seperable Hilbert space) for things to work out~\citep{da_prato_stochastic_2014,balan_gentle_2018}, and I do not know whether our domain is nice enough. For example, it is not immediately clear what conditions on $\mathcal{A}^{\pm}(x | \phi)$ are needed for \eqref{functional_langevin} to even admit a solution. Nevertheless, we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, and under certain assumptions, equations like \eqref{nD_eqn_for_frequencies} have been proved to exist.

\section{The infinite population limit}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `PDE models' discussed in studies of Adaptive Diversification~\citep{doebeli_adaptive_2011}. Equation \eqref{deterministic_traj} is also the basic equation of `oligomorphic dynamics'~\citep{sasaki_oligomorphic_2011, lion_multimorph_2022} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Models of the form \eqref{deterministic_traj} are also used to study intraspecific trait variation in community ecology~\citep{nordbotten_dynamics_2020}. A prominent recent example is the `trait space equations' of~\citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics. As before, I assume that the birth and death functions take the form:

@@

As in chapter \ref{chap_BD}, $ Q(x|\psi)$ describes birth due to mutations and $\mu \geq 0$ is a constant mutation rate. The functions $b^{\textrm{(ind)}}(x|\psi)$ and $d^{\textrm{(ind)}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ individuals in a population $\psi$. These functions could in principle model several ecological factors. For example,  $b^{\textrm{(ind)}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rates in the asexual case, and $d^{\textrm{(ind)}}(x|\psi)$ may model death due to intraspecific competition for resources. Note that the definition of the mean value \eqref{nD_mean} of a type level quantity $f(x)$ now becomes
\overline{f}(t) = \int\limits_{\mathcal{T}}f(x)p(x,t)dx
Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain

@@

\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu Q(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu Q(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} Q(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} Q(y|\psi)dy\right)

@@

\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} Q(y|\psi)dy\right]}

@@

\frac{d \overline{f}}{dt} &= \int\limits_{\mathcal{T}}f(x,t)w(x|\psi)p(x,t)dx - \overline{w}(t)\int\limits_{\mathcal{T}}f(x,t)p(x,t)dx\nonumber\\
&+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} Q(y|\psi)dy\right]dx\nonumber\\
&= \overline{fw} - \overline{w}\cdot\overline{f} + \frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}}f(x,t)\left[Q(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} Q(y|\psi)dy\right]dx\label{intermediate_for_cts_price}
